---
title: "eDNA CPUE Comparison"
author: "Sam Matthews"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## 1. Setup and Global Functions

Load the necessary libraries and create helper functions that will be used for modeling, evaluation, and plotting throughout the analysis.

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(readxl)
library(fuzzyjoin)
library(broom)
library(MASS)
library(glmmTMB)
library(mgcv)
library(tidyr)
library(patchwork)
library(rsample)
library(purrr)
library(tibble)
library(viridis)
library(scales)

# Setup a clean plotting theme
theme_set(theme_bw() + theme(panel.grid.minor = element_blank()))
```

```{r helper-functions}
# Formatting labels for linear models
lm_label <- function(model, digits_r2 = 2, digits_p = 3, digits_rmse = 2) {
    g <- glance(model)
    r2 <- g$r.squared
    p <- g$p.value
    if (is.na(p)) {
        fs <- summary(model)$fstatistic
        p <- pf(fs[1], fs[2], fs[3], lower.tail = FALSE)
    }
    rmse <- sqrt(mean(residuals(model)^2))
    p_txt <- if (p < 0.001) "< 0.001" else formatC(p, format = "f", digits = digits_p)
    sprintf(paste0("R^2 = %.", digits_r2, "f\n", "p = %s\n", "RMSE = %.", digits_rmse, "f"), r2, p_txt, rmse)
}

# Confusion matrices and Evaluation Metrics
make_confusion <- function(dat, perc_thresh, cpue_thresh) {
    classified <- dat %>%
        mutate(
            pred   = if_else(perc_pos >= perc_thresh, "Pred +", "Pred -"),
            actual = if_else(cpue >= cpue_thresh, "Actual +", "Actual -")
        )

    cm <- classified %>%
        count(actual, pred, .drop = FALSE) %>%
        complete(
            actual = c("Actual +", "Actual -"),
            pred = c("Pred +", "Pred -"),
            fill = list(n = 0)
        ) %>%
        group_by(actual) %>%
        mutate(row_prop = n / sum(n)) %>%
        ungroup() %>%
        mutate(label = paste0(n, "\n", scales::percent(row_prop, accuracy = 0.1)))

    TP <- cm$n[cm$actual == "Actual +" & cm$pred == "Pred +"]
    FP <- cm$n[cm$actual == "Actual -" & cm$pred == "Pred +"]
    TN <- cm$n[cm$actual == "Actual -" & cm$pred == "Pred -"]
    FN <- cm$n[cm$actual == "Actual +" & cm$pred == "Pred -"]

    acc <- (TP + TN) / (TP + TN + FP + FN + 1e-9)
    sens <- TP / (TP + FN + 1e-9) # Recall / TPR
    spec <- TN / (TN + FP + 1e-9)
    prec <- TP / (TP + FP + 1e-9) # PPV
    f1 <- 2 * prec * sens / (prec + sens + 1e-9)

    list(
        cm = cm,
        metrics = tibble(
            perc_thresh = perc_thresh, cpue_thresh = cpue_thresh, Accuracy = acc,
            Sensitivity_TPR = sens, Specificity_TNR = spec, Precision_PPV = prec, F1 = f1
        )
    )
}

plot_confusion <- function(cm_obj, title = "Confusion matrix") {
    ggplot(cm_obj$cm, aes(x = pred, y = actual, fill = row_prop)) +
        geom_tile(color = "grey85", linewidth = 0.4) +
        geom_text(aes(label = label), size = 5) +
        scale_fill_viridis_c(option = "mako", direction = -1, begin = 0.25, name = "Count Rate") +
        labs(x = "Prediction from eDNA (% pos)", y = "Ground truth from CPUE", title = title) +
        coord_fixed()
}

metrics_for <- function(p_thr, c_thr, x) {
    pred <- x$perc_pos >= p_thr
    actual <- x$cpue >= c_thr
    TP <- sum(pred & actual)
    FP <- sum(pred & !actual)
    TN <- sum(!pred & !actual)
    FN <- sum(!pred & actual)
    prec <- TP / (TP + FP + 1e-9)
    rec <- TP / (TP + FN + 1e-9)
    f1 <- 2 * prec * rec / (prec + rec + 1e-9)
    acc <- (TP + TN) / (TP + TN + FP + FN + 1e-9)
    c(F1 = f1, Accuracy = acc, Precision = prec, Recall = rec)
}
```

## 2. Data Loading and Aggregation

Read raw CPUE (cull) and eDNA data.

**Analytical Fixes Implemented:**
- Swapped CSV reading for native Excel reading via `readxl`.
- Removed explicit character-to-date formatting strings (`"%d/%m/%Y"`) that fail on native Excel timestamps, substituting to rely on `as.Date` acting structurally on numeric-date representations.
- Extracted and normalized column names with explicit checks.

```{r Load Data}
# Load Cull data
cull.dat <- read_excel("data/260201_COTS-Cull-Data-Ewels.xlsx", sheet = "Cull")

# Load eDNA data
edna.dat <- read_excel("data/eDNA data_ALL_20260225.xlsx", sheet = "eDNA_data_ALL")

# Clean Cull Data
cull <- cull.dat %>%
    rename(Reef = ReefName) %>%
    mutate(date_cull = as.Date(SurveyDate))

# Aggregate eDNA data
edna_agg <- edna.dat %>%
    filter(!is.na(Year)) %>%
    rename(Collection.organisation = `Collection organisation`) %>%
    mutate(
        Reef = ReefName,
        Collection.org = ifelse(Collection.organisation == "AIMS", "AIMS", "Other"),
        date_edna = as.Date(Date), # Excel supplies native dates
        Conc_mean = as.numeric(Conc_mean)
    ) %>%
    arrange(Reef, Year, date_edna) %>%
    group_by(Reef, Collection.org, Year) %>%
    # cluster rows into groups where consecutive samples are ≤ 7 days apart
    mutate(
        grp = cumsum(
            if_else(
                is.na(lag(date_edna)) | as.numeric(date_edna - lag(date_edna)) > 7,
                1L, 0L
            )
        )
    ) %>%
    ungroup() %>%
    group_by(Reef, Collection.org, Year, grp) %>%
    summarise(
        date_edna = min(date_edna),
        conc_mean = mean(Conc_mean, na.rm = TRUE),
        perc_pos  = mean(LOD_sample_positive, na.rm = TRUE) * 100,
        n_samples = n(),
        .groups   = "drop"
    )
```

## 3. Fuzzy Temporal Joining

Merge datasets geographically bounding matches within varying time intervals.

### 3.1 Matches within ± 6 months (183 days)

```{r join-any6}
win_days <- 183

same6_any <- fuzzy_inner_join(
    cull, edna_agg,
    by = c("Reef" = "Reef", "date_cull" = "date_edna"),
    match_fun = list(
        `==`,
        function(cull_date, edna_date) {
            abs(as.numeric(cull_date - edna_date)) <= win_days
        }
    )
) %>%
    mutate(diff_days = as.numeric(date_cull - date_edna), Reef = Reef.x)

reef_any <- same6_any %>%
    group_by(Reef, Collection.org) %>%
    summarise(
        total_cots      = sum(Cohort1 + Cohort2 + Cohort3 + Cohort4, na.rm = TRUE),
        total_bottom    = sum(Bottomtime, na.rm = TRUE),
        cpue_reef       = total_cots / total_bottom,
        perc_pos_reef   = mean(perc_pos, na.rm = TRUE),
        conc_mean_reef  = mean(conc_mean, na.rm = TRUE),
        n_matches       = n(),
        .groups         = "drop"
    )
```

### 3.2 Matches with eDNA Prior to Cull (3, 6, and 12 Months)

This represents the primary actionability condition: *Did a prior eDNA signal capture an ongoing or succeeding COTS outbreak?* We evaluate three horizons to observe signal degradation.

```{r join-temp-horizons}
get_prior_cohort <- function(culls, ednas, win_days, label) {
    res <- fuzzy_inner_join(
        culls, ednas,
        by = c("Reef" = "Reef", "date_cull" = "date_edna"),
        match_fun = list(
            `==`,
            function(cull_date, edna_date) {
                diff <- as.numeric(cull_date - edna_date)
                diff >= 0 & diff <= win_days
            }
        )
    ) %>% mutate(diff_days = as.numeric(date_cull - date_edna), Reef = Reef.x)

    res %>%
        group_by(Reef, Collection.org, Year, grp) %>%
        summarise(
            horizon         = label,
            total_cots      = sum(Cohort1 + Cohort2 + Cohort3 + Cohort4, na.rm = TRUE),
            total_bottom    = sum(Bottomtime, na.rm = TRUE),
            cpue_reef       = total_cots / total_bottom,
            perc_pos_reef   = mean(perc_pos, na.rm = TRUE),
            conc_mean_reef  = mean(conc_mean, na.rm = TRUE),
            n_matches       = n(),
            .groups         = "drop"
        )
}

reef_prior_3m <- get_prior_cohort(cull, edna_agg, 91, "3 Months")
reef_prior_6m <- get_prior_cohort(cull, edna_agg, 183, "6 Months")
reef_prior_12m <- get_prior_cohort(cull, edna_agg, 365, "12 Months")

reef_before <- bind_rows(reef_prior_3m, reef_prior_6m, reef_prior_12m) %>%
    mutate(horizon = factor(horizon, levels = c("3 Months", "6 Months", "12 Months")))
```

## 4. Initial Linear Comparisons

```{r exploratory-plots, fig.width=10, fig.height=5}
p1 <- ggplot(reef_before, aes(x = perc_pos_reef, y = cpue_reef, color = horizon)) +
    geom_point(alpha = 0.5, size = 2) +
    geom_smooth(method = "lm", se = TRUE) +
    geom_hline(yintercept = 0.04, linetype = "dashed", color = "purple") +
    labs(x = "% eDNA positive", y = "CPUE", title = "Prior eDNA (% Pos) vs CPUE") +
    facet_wrap(~horizon)

p2 <- ggplot(reef_before, aes(x = log1p(conc_mean_reef), y = cpue_reef, color = horizon)) +
    geom_point(alpha = 0.5, size = 2) +
    geom_smooth(method = "lm", se = TRUE) +
    geom_hline(yintercept = 0.04, linetype = "dashed", color = "purple") +
    labs(x = "log1p(Mean Concentration)", y = "CPUE", title = "Prior eDNA (Conc) vs CPUE") +
    facet_wrap(~horizon)

p1 / p2
```

## 5. Generalized Linear Mixed-Effects Models (GLMMs)

We model discrete outcomes (observed COTS numbers) strictly to better encompass count structure offsets driven by varying bottom time. We filter extreme `conc_mean_reef >= 5000` to constrain highly anomalous impacts on variance gradients.

```{r glmm-setup}
# Filter outliers systematically at modeling onset
dat_glmm <- reef_before %>%
    mutate(
        conc_t    = log1p(conc_mean_reef),
        obs_cpue  = total_cots / total_bottom
    ) %>%
    filter(
        !is.na(total_cots), !is.na(total_bottom), total_bottom > 0,
        !is.na(perc_pos_reef), !is.na(conc_t), !is.na(Reef),
        conc_mean_reef < 5000
    ) %>%
    mutate(Reef = factor(Reef)) %>%
    droplevels()

effort_ref <- median(dat_glmm$total_bottom, na.rm = TRUE)
```

### 5.1 Negative Binomial (% Positive)

```{r glmm-perc, fig.width=10, fig.height=4}
fit_horizon_glmm <- function(hor_val) {
    dat_sub <- dat_glmm %>% filter(horizon == hor_val)
    if (nrow(dat_sub) == 0) {
        return(NULL)
    }

    m_nb <- glmmTMB(
        total_cots ~ perc_pos_reef + offset(log(total_bottom)) + (1 | Reef),
        family = nbinom2(),
        data = dat_sub,
        control = glmmTMBControl(optCtrl = list(iter.max = 1e3, eval.max = 1e3))
    )

    nd <- tibble(
        perc_pos_reef = seq(min(dat_sub$perc_pos_reef), max(dat_sub$perc_pos_reef), length.out = 100),
        total_bottom = median(dat_sub$total_bottom, na.rm = TRUE),
        Reef = NA,
        horizon = hor_val
    )

    pred <- predict(m_nb, newdata = nd, type = "link", se.fit = TRUE, re.form = NA)
    nd %>%
        mutate(
            fit_cpue  = exp(pred$fit) / total_bottom,
            lcl_cpue  = exp(pred$fit - 1.96 * pred$se.fit) / total_bottom,
            ucl_cpue  = exp(pred$fit + 1.96 * pred$se.fit) / total_bottom
        )
}

horizons <- levels(dat_glmm$horizon)
nd_all <- map_dfr(horizons, fit_horizon_glmm) %>%
    mutate(horizon = factor(horizon, levels = horizons))

ggplot(dat_glmm, aes(x = perc_pos_reef, y = obs_cpue)) +
    geom_point(alpha = 0.5) +
    geom_hline(yintercept = 0.04, linetype = "dashed", color = "purple") +
    geom_ribbon(data = nd_all, aes(x = perc_pos_reef, ymin = lcl_cpue, ymax = ucl_cpue, fill = horizon), alpha = 0.3, inherit.aes = FALSE) +
    geom_line(data = nd_all, aes(x = perc_pos_reef, y = fit_cpue, color = horizon), linewidth = 1, inherit.aes = FALSE) +
    facet_wrap(~horizon) +
    labs(x = "% eDNA positive", y = "CPUE", title = "NegBin GLM Offset Modeling: % Pos Degradation") +
    theme(legend.position = "none")
```

### 5.2 Negative Binomial (Both Predictors w/ Interaction)

```{r glmm-interaction}
m_nb_base <- glmmTMB(
    total_cots ~ perc_pos_reef + offset(log(total_bottom)) + (1 | Reef),
    family = nbinom2(),
    data = dat_glmm
)

m_nb_both <- glmmTMB(
    total_cots ~ perc_pos_reef * conc_t + offset(log(total_bottom)) + (1 | Reef),
    family = nbinom2(),
    data = dat_glmm
)

# Interaction Anova Validation
anova(m_nb_base, m_nb_both)
```

## 6. Generalized Additive Models (GAM)

Comparing linear assumptions against structured splines.

```{r gam-perc}
g_perc <- gam(
    total_cots ~ s(perc_pos_reef, k = 3) + s(Reef, bs = "re") + offset(log(total_bottom)),
    family = nb(), method = "REML",
    data = dat_glmm
)

nd_gam <- tibble(
    perc_pos_reef = seq(min(dat_glmm$perc_pos_reef), max(dat_glmm$perc_pos_reef), length.out = 200),
    total_bottom = effort_ref,
    Reef = dat_glmm$Reef[1]
)

pp_perc <- predict(g_perc, newdata = nd_gam, type = "link", se.fit = TRUE, exclude = "s(Reef)")
nd_gam <- nd_gam %>%
    mutate(
        fit_cpue  = exp(pp_perc$fit) / total_bottom,
        lcl_cpue  = exp(pp_perc$fit - 1.96 * pp_perc$se.fit) / total_bottom,
        ucl_cpue  = exp(pp_perc$fit + 1.96 * pp_perc$se.fit) / total_bottom
    )

ggplot(dat_glmm, aes(x = perc_pos_reef, y = obs_cpue)) +
    geom_point(alpha = 0.7) +
    geom_hline(yintercept = 0.04, linetype = "dashed", color = "purple") +
    geom_ribbon(data = nd_gam, aes(x = perc_pos_reef, ymin = lcl_cpue, ymax = ucl_cpue), alpha = 0.3, fill = "skyblue", inherit.aes = FALSE) +
    geom_line(data = nd_gam, aes(x = perc_pos_reef, y = fit_cpue), linewidth = 1, color = "black", inherit.aes = FALSE) +
    labs(title = "GAM Spline NegBin Estimate: % Pos")
```

## 7. Confusion Matrix Decision Modeling

We iterate over thresholds matching traditional Management logic.

```{r confusion-standard}
# Set CPUE Management Threshold
cpue_thresh <- 0.04

# Baseline uses 6 Months max window; Aggregating strictly by Reef
dat_evt <- dat_glmm %>%
    filter(horizon == "6 Months") %>%
    group_by(Reef, Collection.org) %>%
    summarise(
        counts = sum(total_cots, na.rm = TRUE),
        total_bottom = sum(total_bottom, na.rm = TRUE),
        perc_pos = mean(perc_pos_reef, na.rm = TRUE),
        conc_mean_reef = mean(conc_mean_reef, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    mutate(
        reef = Reef,
        cpue = counts / total_bottom,
        conc_t = log1p(conc_mean_reef)
    ) %>%
    dplyr::select(reef, Collection.org, perc_pos, conc_t, conc_mean_reef, cpue)

res_all <- make_confusion(dat_evt, perc_thresh = 48, cpue_thresh = cpue_thresh)
plot_confusion(res_all, title = sprintf("Full Data Matrix (eDNA ≥ 48%%, CPUE ≥ %.3f)", cpue_thresh))
```

## 8. Threshold Selection via Cross Validation

We utilize out-of-fold `rsample` stratification testing to derive optimal classification rules explicitly guarding F1 recall bounds under variable density environments.

```{r cross-validation, fig.width=10, fig.height=8}
# 1) Evaluate on a grid of %pos AND cpue thresholds (Heatmap)
perc_grid <- seq(0, 100, by = 2)
cpue_grid <- seq(0, quantile(dat_evt$cpue, 0.99, na.rm = TRUE), length.out = 40)

grid <- tidyr::expand_grid(perc_thresh = perc_grid, cpue_thresh = cpue_grid) %>%
    mutate(
        out = purrr::pmap(., ~ metrics_for(..1, ..2, dat_evt)),
        F1 = vapply(out, `[[`, numeric(1), "F1"),
        Accuracy = vapply(out, `[[`, numeric(1), "Accuracy")
    ) %>%
    dplyr::select(-out)

# F1 Heatmap
p_hm <- ggplot(grid, aes(x = perc_thresh, y = cpue_thresh, fill = F1)) +
    geom_raster() +
    geom_contour(aes(z = F1), breaks = seq(0.2, 1, by = 0.2), colour = "black", linewidth = 0.3, alpha = 0.7) +
    scale_fill_viridis_c(limits = c(0, 1)) +
    labs(x = "eDNA % positive threshold", y = "CPUE threshold", fill = "F1", title = "F1 across %pos thresholds with iso-F1 contours")

# 2) Cross validate across multiple CPUE levels
cpue_levels <- c(0.01, 0.02, 0.04, 0.06, 0.08, 0.10)
k <- 5
R <- 100 # Reduced from 200 for rendering speed
set.seed(1)

run_cv_for_cpue <- function(dat_evt, cpue_thr, perc_grid, k = 5, R = 100) {
    dat2 <- dat_evt %>% mutate(actual = if (cpue_thr == 0) (cpue > 0) else (cpue >= cpue_thr))
    strat_ok <- length(unique(dat2$actual)) > 1
    folds <- if (strat_ok) vfold_cv(dat2, v = k, repeats = R, strata = actual) else vfold_cv(dat2, v = k, repeats = R)

    cv_long <- folds %>%
        mutate(assess = map(splits, assessment)) %>%
        dplyr::select(id, id2, assess) %>%
        tidyr::expand_grid(perc_thresh = perc_grid) %>%
        mutate(
            out = pmap(list(perc_thresh, assess), \(p, d) metrics_for(p, cpue_thr, d)),
            F1 = vapply(out, `[[`, numeric(1), "F1"),
            Accuracy = vapply(out, `[[`, numeric(1), "Accuracy"),
            Precision = vapply(out, `[[`, numeric(1), "Precision"),
            Recall = vapply(out, `[[`, numeric(1), "Recall")
        ) %>%
        dplyr::select(-out, -assess)

    cv_sum <- cv_long %>%
        group_by(perc_thresh) %>%
        summarise(
            F1_mean = mean(F1, na.rm = TRUE), F1_sd = sd(F1, na.rm = TRUE), F1_se = F1_sd / sqrt(n()), .groups = "drop"
        )

    best <- cv_sum %>% slice_max(F1_mean, n = 1, with_ties = FALSE)
    p_star <- best$perc_thresh

    perf_split <- folds %>%
        mutate(
            assess = map(splits, assessment),
            m = map(assess, ~ {
                out <- metrics_for(p_star, cpue_thr, .x)
                tibble(F1 = out["F1"], Accuracy = out["Accuracy"], Precision = out["Precision"], Recall = out["Recall"])
            })
        ) %>%
        dplyr::select(m) %>%
        unnest(m)

    perf_sum <- perf_split %>%
        summarise(
            cpue_thr = cpue_thr, perc_star = p_star,
            F1_mean = mean(F1), F1_sd = sd(F1), F1_se = F1_sd / sqrt(n()),
            Acc_mean = mean(Accuracy), Acc_sd = sd(Accuracy), Acc_se = Acc_sd / sqrt(n()),
            Prec_mean = mean(Precision), Prec_sd = sd(Precision), Prec_se = Prec_sd / sqrt(n()),
            Rec_mean = mean(Recall), Rec_sd = sd(Recall), Rec_se = Rec_sd / sqrt(n()),
            .groups = "drop"
        )

    cm_full <- make_confusion(dat2 %>% dplyr::select(perc_pos, cpue), p_star, cpue_thr)

    list(cpue_thr = cpue_thr, perc_star = p_star, cv_sum = cv_sum, perf_sum = perf_sum, cm_full = cm_full)
}

results <- map(cpue_levels, ~ run_cv_for_cpue(dat_evt, .x, perc_grid, k = k, R = R))

summary_table <- map_dfr(results, "perf_sum") %>%
    mutate(
        F1_mean_se   = sprintf("%.3f ± %.3f", F1_mean, F1_se),
        Acc_mean_se  = sprintf("%.3f ± %.3f", Acc_mean, Acc_se),
        Prec_mean_se = sprintf("%.3f ± %.3f", Prec_mean, Prec_se),
        Rec_mean_se  = sprintf("%.3f ± %.3f", Rec_mean, Rec_se)
    ) %>%
    dplyr::select(cpue_thr, perc_star, F1_mean_se, Acc_mean_se, Prec_mean_se, Rec_mean_se)

# Show the optimal threshold summary
summary_table

# Plot multi-panel confusion matrices
cm_all <- map_dfr(results, function(r) {
    r$cm_full$cm %>% mutate(cpue_thr = r$cpue_thr, perc_star = r$perc_star)
}) %>% mutate(panel = paste0("CPUE ≥ ", cpue_thr, "\n%pos* = ", perc_star))

p_cm <- ggplot(cm_all, aes(x = pred, y = actual, fill = row_prop)) +
    geom_tile(color = "grey85", linewidth = 0.4) +
    geom_text(aes(label = label), size = 4, color = "black") +
    facet_wrap(~panel, ncol = 3) +
    scale_fill_viridis_c(option = "mako", direction = -1, begin = 0.25, name = "Count Rate") +
    labs(x = "Prediction from eDNA (% pos)", y = "Ground truth from CPUE", title = "Confusion matrices with CV-tuned thresholds") +
    coord_fixed()

p_hm
p_cm
```

### 8.1 Horizon-Specific Thresholding (CPUE ≥ 0.02)

To evaluate eDNA signal durability, we cross-validate optimal `% Pos` thresholds iteratively across our three temporal windows (`3 Months`, `6 Months`, `12 Months`) targeting the management CPUE constraint of `0.02`.

```{r cv-horizon, fig.width=10, fig.height=4}
cpue_target <- 0.02

# Ensure we aggregate strictly by Reef to prevent Site-level contamination
dat_evt_hor <- dat_glmm %>%
    group_by(Reef, Collection.org, horizon) %>%
    summarise(
        counts = sum(total_cots, na.rm = TRUE),
        total_bottom = sum(total_bottom, na.rm = TRUE),
        perc_pos = mean(perc_pos_reef, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    mutate(
        reef = Reef,
        cpue = counts / total_bottom
    ) %>%
    dplyr::select(reef, Collection.org, perc_pos, cpue, horizon)

run_cv_for_horizon <- function(dat_full, hor_val, cpue_thr, perc_grid, k = 5, R = 100) {
    dat_sub <- dat_full %>% filter(horizon == hor_val)
    res <- run_cv_for_cpue(dat_sub, cpue_thr, perc_grid, k = k, R = R)
    res$perf_sum <- res$perf_sum %>% mutate(horizon = hor_val)
    res$cm_full$cm <- res$cm_full$cm %>% mutate(horizon = hor_val)
    res
}

horizons <- levels(dat_evt_hor$horizon)
results_hor <- map(horizons, ~ run_cv_for_horizon(dat_evt_hor, .x, cpue_target, perc_grid, k = k, R = R))

summary_hor <- map_dfr(results_hor, "perf_sum") %>%
    mutate(
        F1_mean_se   = sprintf("%.3f ± %.3f", F1_mean, F1_se),
        Acc_mean_se  = sprintf("%.3f ± %.3f", Acc_mean, Acc_se),
        Prec_mean_se = sprintf("%.3f ± %.3f", Prec_mean, Prec_se),
        Rec_mean_se  = sprintf("%.3f ± %.3f", Rec_mean, Rec_se)
    ) %>%
    dplyr::select(horizon, cpue_thr, perc_star, F1_mean_se, Acc_mean_se, Prec_mean_se, Rec_mean_se)

summary_hor

cm_all_hor <- map_dfr(results_hor, function(r) {
    r$cm_full$cm %>% mutate(cpue_thr = r$cpue_thr, perc_star = r$perc_star)
}) %>% mutate(panel = factor(paste0(horizon, "\n%pos* = ", perc_star), levels = paste0(levels(dat_evt_hor$horizon), "\n%pos* = ", map_dbl(results_hor, "perc_star"))))

p_cm_hor <- ggplot(cm_all_hor, aes(x = pred, y = actual, fill = row_prop)) +
    geom_tile(color = "grey85", linewidth = 0.4) +
    geom_text(aes(label = label), size = 4, color = "black") +
    facet_wrap(~panel, ncol = 3) +
    scale_fill_viridis_c(option = "mako", direction = -1, begin = 0.25, name = "Count Rate") +
    labs(x = "Prediction from eDNA (% pos)", y = "Ground truth from CPUE", title = sprintf("Confusion Matrices across Horizons (CPUE ≥ %.2f)", cpue_target)) +
    coord_fixed()

p_cm_hor
```

## 9. Multivariate Rule Search

Combining `% Pos` AND `% Conc` conditions under 2D constraints. By modeling concentration in log-transformed space (`conc_t`), we fix skewed grid sparsity and resolve the F1 heatmaps dropping out continuously.

```{r 2D-search, fig.width=10, fig.height=8}
metrics_for_both <- function(p_thr, c_thr, cpue_thr, rule = c("or", "and"), x) {
    rule <- match.arg(rule)
    pred <- if (rule == "or") {
        (x$perc_pos >= p_thr) | (x$conc_t >= c_thr)
    } else {
        (x$perc_pos >= p_thr) & (x$conc_t >= c_thr)
    }
    actual <- x$cpue >= cpue_thr
    TP <- sum(pred & actual)
    FP <- sum(pred & !actual)
    TN <- sum(!pred & !actual)
    FN <- sum(!pred & actual)
    prec <- TP / (TP + FP + 1e-9)
    rec <- TP / (TP + FN + 1e-9)
    f1 <- 2 * prec * rec / (prec + rec + 1e-9)
    acc <- (TP + TN) / (TP + TN + FP + FN + 1e-9)
    c(F1 = f1, Accuracy = acc, Precision = prec, Recall = rec)
}

# Tune on transformed concentration to avoid skewness clustering 0s all the way across
conc_t_max <- quantile(dat_evt$conc_t, 0.99, na.rm = TRUE)
conc_grid_t <- seq(min(dat_evt$conc_t, na.rm = TRUE), conc_t_max, length.out = 30)

# Evaluate OR and AND grids over 2D thresholds for a baseline cpue
cpue_star <- 0.04

grid_both_and <- tidyr::expand_grid(perc_thresh = seq(0, 100, by = 2), conc_thresh_t = conc_grid_t) %>%
    mutate(
        out = purrr::pmap(., ~ metrics_for_both(..1, ..2, cpue_star, "and", dat_evt)),
        F1 = vapply(out, `[[`, numeric(1), "F1")
    )

p_2dhm <- ggplot(grid_both_and, aes(x = perc_thresh, y = expm1(conc_thresh_t), fill = F1)) +
    geom_raster() +
    geom_contour(aes(z = F1), breaks = seq(0.2, 0.9, by = 0.2), colour = "black", linewidth = 0.3) +
    scale_fill_viridis_c(limits = c(0, 1)) +
    scale_y_continuous(trans = scales::pseudo_log_trans(sigma = 1)) +
    labs(x = "% eDNA Positive Threshold", y = "Concentration Threshold (pseudo-log)", title = sprintf("2D Boolean Search: AND Rule F1 (CPUE ≥ %.3f)", cpue_star))


# Full 2D Loop over CPUE levels
run_cv_2d <- function(dat, cpue_thr, perc_grid, conc_grid_t, rule = "and") {
    best_pair <- function(df, cpue_thr) {
        actual <- df$cpue >= cpue_thr
        P <- length(perc_grid)
        C <- length(conc_grid_t)
        perc_ge <- outer(df$perc_pos, perc_grid, `>=`)
        conc_ge <- outer(df$conc_t, conc_grid_t, `>=`)
        f1_mat <- matrix(NA_real_, nrow = P, ncol = C)
        for (pi in seq_len(P)) {
            pred_mat <- if (rule == "and") conc_ge & perc_ge[, pi] else conc_ge | perc_ge[, pi]
            TP <- colSums(pred_mat & actual)
            FP <- colSums(pred_mat & !actual)
            FN <- colSums(!pred_mat & actual)
            prec <- TP / (TP + FP + 1e-9)
            rec <- TP / (TP + FN + 1e-9)
            f1_mat[pi, ] <- 2 * prec * rec / (prec + rec + 1e-9)
        }
        max_f1 <- max(f1_mat, na.rm = TRUE)
        idx <- which(f1_mat == max_f1, arr.ind = TRUE)
        idx <- idx[order(idx[, 1], idx[, 2], decreasing = TRUE), , drop = FALSE][1, ]
        list(perc_star = perc_grid[idx[1]], conc_t_star = conc_grid_t[idx[2]])
    }

    dat2 <- dat %>% mutate(actual = cpue >= cpue_thr)
    if (length(unique(dat2$actual)) < 2) {
        return(NULL)
    }

    folds <- vfold_cv(dat2, v = k, repeats = R, strata = actual)

    split_res <- map_dfr(folds$splits, function(spl) {
        tr <- analysis(spl)
        te <- assessment(spl)
        best <- best_pair(tr, cpue_thr)
        met <- metrics_for_both(best$perc_star, best$conc_t_star, cpue_thr, rule, te)
        tibble(perc_star = best$perc_star, conc_t_star = best$conc_t_star, conc_mean_star = pmax(expm1(best$conc_t_star), 0)) %>%
            bind_cols(tibble(F1 = met["F1"], Accuracy = met["Accuracy"], Precision = met["Precision"], Recall = met["Recall"]))
    })

    thr <- split_res %>% summarise(perc_star = median(perc_star), conc_t_star = median(conc_t_star), conc_mean_star = median(conc_mean_star), .groups = "drop")

    perf <- split_res %>%
        summarise(
            cpue_thr = cpue_thr, perc_star = thr$perc_star, conc_mean_star = thr$conc_mean_star,
            F1_mean = mean(F1), F1_se = sd(F1) / sqrt(n()),
            Acc_mean = mean(Accuracy), Acc_se = sd(Accuracy) / sqrt(n()),
            Prec_mean = mean(Precision), Prec_se = sd(Precision) / sqrt(n()),
            Rec_mean = mean(Recall), Rec_se = sd(Recall) / sqrt(n()), .groups = "drop"
        )

    list(thr = thr, perf = perf)
}

results_2d <- map(cpue_levels, ~ run_cv_2d(dat_evt, .x, seq(0, 100, by = 5), conc_grid_t, rule = "and")) # %>% discard(is.null)

summary_table_2d <- map_dfr(results_2d, "perf") %>%
    mutate(
        F1_mean_se   = sprintf("%.3f ± %.3f", F1_mean, F1_se),
        Acc_mean_se  = sprintf("%.3f ± %.3f", Acc_mean, Acc_se),
        Prec_mean_se = sprintf("%.3f ± %.3f", Prec_mean, Prec_se),
        Rec_mean_se  = sprintf("%.3f ± %.3f", Rec_mean, Rec_se)
    ) %>%
    dplyr::select(cpue_thr, perc_star, conc_mean_star, F1_mean_se, Acc_mean_se, Prec_mean_se, Rec_mean_se)

# Show operational 2D thresholds
summary_table_2d

# Plot 2D Multi-CM
make_confusion_2d <- function(dat, perc_star, conc_t_star, cpue_thr, rule = "and") {
    classified <- dat %>%
        mutate(
            pred = if_else(perc_pos >= perc_star & conc_t >= conc_t_star, "Pred +", "Pred -"),
            actual = if_else(cpue >= cpue_thr, "Actual +", "Actual -")
        )
    cm <- classified %>%
        count(actual, pred, .drop = FALSE) %>%
        complete(actual = c("Actual +", "Actual -"), pred = c("Pred +", "Pred -"), fill = list(n = 0)) %>%
        group_by(actual) %>%
        mutate(row_prop = n / sum(n)) %>%
        ungroup() %>%
        mutate(label = paste0(n, "\n", scales::percent(row_prop, accuracy = 0.1)))
    list(cm = cm)
}

thr_df <- map_dfr(results_2d, "thr") %>% mutate(cpue_thr = sapply(results_2d, function(x) x$perf$cpue_thr))
cm_all_2d <- pmap_dfr(list(thr_df$cpue_thr, thr_df$perc_star, thr_df$conc_t_star), function(cp, p, c) {
    obj <- make_confusion_2d(dat_evt, p, c, cp)
    obj$cm %>% mutate(panel = paste0("CPUE ≥ ", cp, "\n%pos*=", round(p, 1), "  conc*≈", round(pmax(expm1(c), 0), 1)))
})

p_2dcm <- ggplot(cm_all_2d, aes(pred, actual, fill = row_prop)) +
    geom_tile(color = "grey85", linewidth = 0.4) +
    geom_text(aes(label = label), size = 4, color = "black") +
    facet_wrap(~panel, ncol = 3) +
    scale_fill_viridis_c(option = "mako", direction = -1, begin = 0.25, limits = c(0, 1), name = "Row Rate") +
    labs(x = "Prediction (2D rule)", y = "Ground truth (CPUE)", title = "2D AND Rule Confusion Matrices") +
    coord_fixed()

p_2dhm
p_2dcm
```

## 10. Site-Level Spatial Buffer Analysis

To evaluate eDNA's predictive power at granular physical site boundaries rather than arbitrary Reefs, we extract explicit geographic boundaries from both datasets. Cultivating representative site locations mitigates broader environmental noise by strictly evaluating COTS overlap physically clustered against the immediate eDNA sample trace.

Importantly, this spatial test formally evaluates differing sampling regimes present in the program (e.g., `4 sites x 6 samples` vs `3 sites x 12 samples`).

```{r spatial-setup, warning=FALSE, message=FALSE}
library(sf)

# 1. Standardize eDNA sample cohorts by their physical 'Site' level rather than 'Reef' layer
edna_site <- edna.dat %>%
    filter(!is.na(Lat), !is.na(Long), !is.na(Year)) %>%
    mutate(date_edna = as.Date(Date)) %>%
    group_by(ReefName, Site_name, Year) %>%
    summarise(
        Lat = mean(Lat),
        Long = mean(Long),
        date_edna = min(date_edna),
        conc_mean = mean(as.numeric(Conc_mean), na.rm = TRUE),
        perc_pos = mean(LOD_sample_positive, na.rm = TRUE) * 100,
        n_samples = n(),
        .groups = "drop"
    ) %>%
    rename(Reef = ReefName)

# 2. Tag Sampling Regimes based on local density
# (avg_samples per site usually bounds around ~6 or ~12)
regime_df <- edna_site %>%
    group_by(Reef, Year) %>%
    summarise(
        avg_samples = mean(n_samples),
        .groups = "drop"
    ) %>%
    mutate(
        regime = case_when(
            avg_samples >= 10 ~ "3x12 Density",
            TRUE ~ "4x6 Density"
        )
    )

edna_site <- edna_site %>% left_join(regime_df, by = c("Reef", "Year"))

# 3. Project coordinate matrices into native EPSG:3112 (Lambert Conformal Conic for AUS)
edna_sf <- st_as_sf(edna_site, coords = c("Long", "Lat"), crs = 4326) %>% st_transform(3112)
cull_sf <- cull %>%
    filter(!is.na(Longitude), !is.na(Latitude)) %>%
    st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
    st_transform(3112)
```

We generate buffers drawn iteratively radially outward (`200m, 500m, 1km, 2km`), isolate matching Culls occurring inside those rings within a 6-month prior trace window, and test CPUE prediction capabilities over regimes.

```{r spatial-buffers, fig.width=12, fig.height=8, message=FALSE, warning=FALSE}
buffer_radii <- c(200, 500, 1000, 2000)
spatial_results <- list()

for (dist_m in buffer_radii) {
    # Radial expansion around each eDNA point
    edna_buf <- st_buffer(edna_sf, dist = dist_m)

    # Isolate Cull coordinates bounding inside the circle
    intersect_cull <- st_join(edna_buf, cull_sf, join = st_intersects) %>%
        filter(!is.na(date_cull))

    # Constrain to 6 month temporal overlaps
    valid_encounters <- intersect_cull %>%
        mutate(diff_days = as.numeric(date_cull - date_edna)) %>%
        filter(diff_days >= 0 & diff_days <= 183)

    # Collapse cull metadata back onto the respective eDNA Site ID
    site_cpue <- valid_encounters %>%
        st_drop_geometry() %>%
        rename(any_of(c(Reef = "Reef.x", Year = "Year.x"))) %>%
        group_by(Reef, Site_name, Year, regime, perc_pos) %>%
        summarise(
            counts = sum(Cohort1 + Cohort2 + Cohort3 + Cohort4, na.rm = TRUE),
            total_bottom = sum(Bottomtime, na.rm = TRUE),
            .groups = "drop"
        ) %>%
        mutate(
            cpue = counts / total_bottom,
            radius = paste0(dist_m, "m Radius")
        )

    spatial_results[[as.character(dist_m)]] <- site_cpue
}

df_spatial <- bind_rows(spatial_results) %>%
    mutate(radius = factor(radius, levels = paste0(buffer_radii, "m Radius")))

# Run CV Threshold tuning evaluating spatial density bounds 
# We evaluate both 0.04 and 0.02 CPUE Targets, using unified global thresholds per radius.
evaluate_unified_spatial <- function(df, cp) {
    cms_all <- list()
    for (r in levels(df$radius)) {
        sub_df <- df %>% filter(radius == r)
        if (nrow(sub_df) < 5) next
        
        # 1. Calculate Single Global Threshold for the combined Radius
        out <- run_cv_for_cpue(sub_df, cp, perc_grid, k = 5, R = 50)
        if (is.null(out)) next
        
        opt_thresh <- out$perc_star
        
        # 2. Apply that same unified threshold across both Regimes individually
        for (reg in unique(sub_df$regime)) {
            reg_df <- sub_df %>% filter(regime == reg)
            if (nrow(reg_df) > 0) {
                cm_reg <- make_confusion(reg_df, perc_thresh = opt_thresh, cpue_thresh = cp)
                cm_reg$cm <- cm_reg$cm %>% mutate(radius = r, regime = reg, cpue_thr = cp, perc_star = opt_thresh)
                cms_all[[paste0(r, "_", reg)]] <- cm_reg$cm
            }
        }
    }
    
    if (length(cms_all) > 0) {
        df_cm <- bind_rows(cms_all) %>%
            mutate(panel = factor(paste0(regime, "\n", radius, "\nGlobal %pos* = ", perc_star)))
            
        p <- ggplot(df_cm, aes(x = pred, y = actual, fill = row_prop)) +
            geom_tile(color = "grey85", linewidth = 0.4) +
            geom_text(aes(label = label), size = 3, color = "black") +
            facet_wrap(~panel, ncol = 4) +
            scale_fill_viridis_c(option = "mako", direction = -1, begin = 0.25, limits = c(0, 1), name = "Row Rate") +
            labs(
                x = "Prediction from eDNA (% pos)",
                y = "Ground truth from explicit Local Site CPUE",
                title = sprintf("Site-Level Confusion Matrices (CPUE ≥ %.2f) with Unified Radial Thresholds", cp)
            ) +
            coord_fixed()
        return(list(p = p, cm = df_cm))
    }
    return(NULL)
}

res_04 <- evaluate_unified_spatial(df_spatial, 0.04)
if (!is.null(res_04)) {
    p_spatial_04 <- res_04$p
    print(p_spatial_04)
}

res_02 <- evaluate_unified_spatial(df_spatial, 0.02)
if (!is.null(res_02)) {
    p_spatial_02 <- res_02$p
    print(p_spatial_02)
}
```
