---
title: "INLA screen: eDNA signal, distance, and time since sampling"
author: "Sam Matthews"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Purpose

This fast first pass asks a narrower operational question than the full SPDE
benchmark: how well do eDNA percent positive, mean concentration, distance from
the eDNA sampling event, and days since that event explain subsequent COTS cull
counts? Spatial and temporal context enter directly through distance and lag.
All models retain one response per cull dive, use a negative-binomial likelihood
with effort offset, and include a reef random intercept. No calendar-year or
GBR-wide latent spatial field is fitted.

```{r setup}
library(dplyr)
library(ggplot2)
library(readxl)
library(sf)

project_root <- normalizePath(
  if (file.exists("data/eDNA data_ALL_20260528.xlsx")) "." else file.path("..", ".."),
  winslash = "/", mustWork = TRUE
)
devtools::load_all(file.path(project_root, "reefDNA"))
output_dir <- file.path(project_root, "analysis", "inla", "output")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
additive_path <- file.path(output_dir, "distance_time_additive_screen.rds")
interaction_path <- file.path(output_dir, "distance_time_interaction_screen.rds")
prediction_path <- file.path(output_dir, "distance_time_site_predictions.rds")
additive_prediction_path <- file.path(output_dir, "distance_time_additive_site_predictions.rds")
refit_screen <- identical(tolower(Sys.getenv("REEFDNA_REFIT_SCREEN", "false")), "true")
reuse_screen <- !refit_screen && file.exists(additive_path) && file.exists(interaction_path)
set.seed(20260908)
```

## Operational eDNA-event linkage

Each cull is paired with its most recent eligible prior eDNA event on the same
reef and within 2,000 m and 183 days. Replicates within that event supply percent
positive and mean concentration. If multiple events have the same lag, the
nearest site is selected. This rule is fixed before inspecting CPUE.

```{r event-linkage}
stage_a <- readRDS(file.path(output_dir, "stage_a_inla_model.rds"))
raw_links <- readRDS(file.path(output_dir, "space_time_links.rds"))
edna_raw <- read_excel(
  file.path(project_root, "data", "eDNA data_ALL_20260528.xlsx"),
  sheet = "eDNA_data_ALL"
)
edna_obs <- prepare_inla_edna(edna_raw, crs_projected = 3112)

event_benchmark <- summarise_inla_event_benchmark(
  raw_links, edna_obs, stage_a$benchmark
)
stopifnot(
  nrow(event_benchmark) == dplyr::n_distinct(event_benchmark$cull_id),
  min(event_benchmark$edna_lag_days) >= 0,
  max(event_benchmark$edna_lag_days) <= 183,
  max(event_benchmark$edna_distance_m) <= 2000
)
saveRDS(event_benchmark, file.path(output_dir, "event_benchmark_data.rds"))

event_audit <- sf::st_drop_geometry(event_benchmark) %>%
  summarise(
    n_unique_culls = n(),
    n_reefs = n_distinct(Reef),
    n_edna_events_selected = n_distinct(event_id),
    median_replicates_per_event_link = median(edna_n_replicates),
    median_distance_m = median(edna_distance_m),
    max_distance_m = max(edna_distance_m),
    median_lag_days = median(edna_lag_days),
    max_lag_days = max(edna_lag_days)
  )
write.csv(event_audit, file.path(output_dir, "event_linkage_audit.csv"), row.names = FALSE)
event_audit
```

## All additive subsets

The four continuous predictors are transformed and standardized once, then the
same complete observations are used for all 16 additive subsets (including the
reef-only null model). Concentration uses `log1p`; distance uses
`log1p(distance / 200)`; lag uses `log1p(days)`.

```{r additive-screen}
if (reuse_screen) {
  additive_screen <- readRDS(additive_path)
} else {
  additive_screen <- fit_inla_cpue_candidates(
    event_benchmark,
    candidates = build_inla_additive_candidates(),
    include_vessel = FALSE,
    num_threads = "2:1"
  )
  saveRDS(additive_screen, additive_path)
}
write.csv(
  additive_screen$ranking,
  file.path(output_dir, "distance_time_additive_ranking.csv"),
  row.names = FALSE
)
additive_screen$ranking
```

## Hierarchical interaction screen

Interactions are tested only with all corresponding main effects retained. Six
models add one pairwise interaction at a time; a final model includes all six
pairwise interactions. Higher-order interactions are deferred because they are
hard to interpret and are more likely to fit sampling idiosyncrasies.

```{r interaction-screen}
if (reuse_screen) {
  interaction_screen <- readRDS(interaction_path)
} else {
  interaction_screen <- fit_inla_cpue_candidates(
    event_benchmark,
    candidates = build_inla_interaction_candidates(),
    include_vessel = FALSE,
    num_threads = "2:1"
  )
  saveRDS(interaction_screen, interaction_path)
}
write.csv(
  interaction_screen$ranking,
  file.path(output_dir, "distance_time_interaction_ranking.csv"),
  row.names = FALSE
)
interaction_screen$ranking
```

## Combined ranking and supported effects

WAIC is the primary fast ranking criterion; mean log CPO is retained as an
observation-level predictive check. A later `brms` analysis should validate the
shortlist using grouped cross-validation by reef or sampling event rather than
selecting solely on in-sample information criteria.

```{r combined-ranking, fig.width=9, fig.height=7}
combined_ranking <- bind_rows(
  mutate(additive_screen$ranking, family = "additive subset"),
  mutate(interaction_screen$ranking, family = "hierarchical interaction")
) %>%
  arrange(waic) %>%
  mutate(
    delta_waic_all = waic - min(waic),
    rank_all = row_number(),
    model_key = paste(family, model, sep = " :: ")
  )
write.csv(
  combined_ranking,
  file.path(output_dir, "distance_time_combined_ranking.csv"),
  row.names = FALSE
)

p_rank <- combined_ranking %>%
  slice_head(n = 15) %>%
  mutate(model_key = reorder(model_key, delta_waic_all)) %>%
  ggplot(aes(delta_waic_all, model_key, color = family)) +
  geom_point(size = 2.8) +
  geom_vline(xintercept = c(2, 7), linetype = "dashed", color = "grey55") +
  labs(
    x = expression(Delta * "WAIC from best formulation"), y = NULL,
    title = "Fast INLA formulation screen",
    subtitle = "Negative-binomial counts with effort offset and reef iid intercept"
  ) +
  theme_bw()
ggsave(
  file.path(output_dir, "distance_time_model_ranking.png"),
  p_rank, width = 9, height = 7, dpi = 300
)
p_rank
combined_ranking %>% slice_head(n = 15)
```

```{r best-coefficients}
best_row <- combined_ranking %>% slice_head(n = 1)
best_screen <- if (best_row$family == "additive subset") additive_screen else interaction_screen
best_fit <- best_screen$fits[[best_row$model]]
best_coefficients <- tibble::rownames_to_column(
  as.data.frame(best_fit$summary.fixed), "term"
)
best_hyperparameters <- tibble::rownames_to_column(
  as.data.frame(best_fit$summary.hyperpar), "parameter"
)
write.csv(best_coefficients, file.path(output_dir, "distance_time_best_coefficients.csv"), row.names = FALSE)
write.csv(best_hyperparameters, file.path(output_dir, "distance_time_best_hyperparameters.csv"), row.names = FALSE)
best_row
best_coefficients
best_hyperparameters
```

## Site-level distance-by-time predictions

The winning formulation is refitted once with prediction rows for the latest
eDNA event at each modelled site. Scenarios cover the observed primary support:
0, 200, 500, 1,000, and 2,000 m, and 0, 30, 90, and 183 days after sampling.

```{r scenario-predictions}
site_scenarios <- build_inla_distance_time_scenarios(
  edna_obs,
  distances_m = c(0, 200, 500, 1000, 2000),
  lag_days = c(0, 30, 90, 183)
) %>%
  filter(.data$Reef %in% best_screen$reef_levels)

if (reuse_screen && file.exists(prediction_path)) {
  site_predictions <- readRDS(prediction_path)
} else {
  best_prediction <- predict_inla_cpue_screen(
    best_screen,
    model_name = best_row$model,
    newdata = site_scenarios,
    num_threads = "2:1"
  )
  site_predictions <- best_prediction$predictions
  saveRDS(site_predictions, prediction_path)
  saveRDS(best_prediction$fit, file.path(output_dir, "distance_time_prediction_fit.rds"))
}
site_prediction_table <- site_predictions %>%
  st_drop_geometry() %>%
  select(
    scenario_id, Reef, site_name, event_id, date_edna,
    edna_prop_positive, edna_conc_mean,
    edna_distance_m, edna_lag_days,
    cpue_mean, cpue_q025, cpue_q10, cpue_q50, cpue_q90, cpue_q975,
    prob_cpue_gt_002, prob_cpue_gt_004
  )
write.csv(
  site_prediction_table,
  file.path(output_dir, "distance_time_site_predictions.csv"),
  row.names = FALSE
)

site_prediction_table %>%
  filter(grepl("Eyrie", Reef, ignore.case = TRUE)) %>%
  arrange(site_name, edna_lag_days, edna_distance_m)
```

For comparison, the best additive model is predicted over the identical grid.
This is the conservative operational candidate until the interaction surface is
confirmed by grouped out-of-sample validation.

```{r additive-scenario-predictions}
best_additive_name <- additive_screen$ranking$model[[1]]
if (reuse_screen && file.exists(additive_prediction_path)) {
  additive_site_predictions <- readRDS(additive_prediction_path)
} else {
  additive_prediction <- predict_inla_cpue_screen(
    additive_screen,
    model_name = best_additive_name,
    newdata = site_scenarios,
    num_threads = "2:1"
  )
  additive_site_predictions <- additive_prediction$predictions
  saveRDS(additive_site_predictions, additive_prediction_path)
}
additive_site_prediction_table <- additive_site_predictions %>%
  st_drop_geometry() %>%
  select(
    scenario_id, Reef, site_name, event_id, date_edna,
    edna_prop_positive, edna_conc_mean,
    edna_distance_m, edna_lag_days,
    cpue_mean, cpue_q025, cpue_q10, cpue_q50, cpue_q90, cpue_q975,
    prob_cpue_gt_002, prob_cpue_gt_004
  )
write.csv(
  additive_site_prediction_table,
  file.path(output_dir, "distance_time_additive_site_predictions.csv"),
  row.names = FALSE
)
```

```{r scenario-plot, fig.width=11, fig.height=9}
scenario_summary <- bind_rows(
  mutate(site_prediction_table, formulation = "All pairwise interactions (WAIC winner)"),
  mutate(additive_site_prediction_table, formulation = "Best additive: % positive + distance + lag")
) %>%
  group_by(formulation, edna_lag_days, edna_distance_m) %>%
  summarise(
    median_cpue = median(cpue_mean),
    median_q10 = median(cpue_q10),
    median_q90 = median(cpue_q90),
    .groups = "drop"
  )

p_scenarios <- ggplot(
  scenario_summary,
  aes(edna_distance_m, median_cpue)
) +
  geom_ribbon(
    aes(ymin = median_q10, ymax = median_q90),
    fill = "#3182bd", alpha = 0.2
  ) +
  geom_line(color = "#08519c", linewidth = 0.9) +
  geom_point(color = "#08519c", size = 2) +
  geom_hline(yintercept = 0.02, linetype = "dashed", color = "#fdae6b") +
  geom_hline(yintercept = 0.04, linetype = "dashed", color = "#e6550d") +
  scale_x_continuous(breaks = c(0, 200, 500, 1000, 2000)) +
  facet_grid(formulation ~ edna_lag_days, labeller = label_both) +
  labs(
    x = "Distance from eDNA sample (m)",
    y = "Median site-level expected CPUE",
    title = "Lightweight INLA distance-by-time scenarios",
    subtitle = "Panels show days since eDNA sampling; ribbons summarize median site-level 80% intervals"
  ) +
  theme_bw()
ggsave(
  file.path(output_dir, "distance_time_site_predictions.png"),
  p_scenarios, width = 11, height = 9, dpi = 300
)
p_scenarios
```

## Interpretation boundary

This screen estimates association and predictive support, not eDNA transport or
causal decay. The selected eDNA event is an operational anchor, and alternative
nearest-event and 365-day link rules should be sensitivity analyses. The best
few formulations should next be refitted in `brms` with grouped out-of-sample
validation and posterior predictive checks before choosing a final model.
