---
title: "eDNA Cost-Benefit Analysis: Optimizing COTS Control"
author: "Analysis Report"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
# Robustly find reefDNA regardless of render context
pkg_path <- if(dir.exists("reefDNA")) "reefDNA" else if(dir.exists("../reefDNA")) "../reefDNA" else "."
devtools::load_all(pkg_path)
library(knitr)
library(ggplot2)
library(scales)
library(tidyr)
library(rpart)
library(leaflet)
library(sf)
library(htmltools)
library(stringr)

# Read the processed data
combined_data <- read.csv("combined_reef_analysis_v2.csv")
top_20_annual <- read.csv("top_20_annual_wasted_v2.csv")
full_matched_data <- read.csv("full_matched_data_v2.csv")

# Baseline Financial Metrics
cost_per_on_water_day <- 12000
diver_hours_per_day <- 24
cost_per_diver_hour <- cost_per_on_water_day / diver_hours_per_day # $500
program_cost <- 300000
total_reefs_surveyed <- 120
tactical_reefs <- 20
operational_tactical_reefs <- tactical_reefs + 20 # User requested repurposing 20 monitoring reefs for a total of 40 tactically scoped reefs
edna_cost_per_reef <- program_cost / total_reefs_surveyed # $2,500
```

## Executive Summary

This report evaluates the potential return on investment (ROI) of incorporating eDNA surveys into the Crown-of-Thorns Starfish (COTS) control program to safely identify and "close" tactical reefs. By eliminating "ghost culling"—the effort spent culling reefs that lack significant COTS populations—the program could redirect significant vessel and diver time.

This analysis relies on actual historical cull efficiency data matched with eDNA records since 2020.

### Baseline Unit Costs

* **Cost per Diver Hour:** `r dollar(cost_per_diver_hour)`
* **eDNA Program Total Cost:** `r dollar(program_cost)`
* **Assumed Number of Tactical Reefs (per year):** `r operational_tactical_reefs` (Assumes 20 targeted reefs + repurposing 20 monitoring reefs)
* **eDNA Average Cost per Reef:** `r dollar(edna_cost_per_reef)`

---

## Historical eDNA Efficacy Assessment (2020-Present)

Rather than just looking at the worst performing reefs, we have broadened our search to **all reefs that have both eDNA and active culling data since 2020**.

We evaluate eDNA rigorously as a diagnostic test to see what would have happened if we explicitly closed reefs based on statistical combinations.

### Classification Logic & Defining "Ghost Culls"

To rigorously score False Positives and False Negatives, we must define the ultimate ground truth.
*   **The Predictor Rule (Prediction):** A statistical cutoff indicating whether we should "Close" our vessel operations or keep the reef "Open" for diving.
*   **The Ground Truth (Actual Cull Outcome):**
    *   *Current Assumption:* An **Outbreak (Needed)** is formally declared if *any* dive on that reef registered a CPUE $> 0.04$. If a reef met this criteria, diving was validated.
    *   *What is a Ghost Cull?* A **Ghost Cull (False Positive)** is classified when the predictor rule says we should keep the reef open to cull it, but following the initial Manta Tows/scouts, the subsequent Culling Data proves that **zero dives ever exceeded 0.04 CPUE**. In other words, if a reef actually had dives recovering $> 0.04$ CPUE, it is completely exonerated—it is *not* a ghost cull, and our models strictly punish the metric (registering a False Negative Missed Outbreak) if we mistakenly attempted to close it.

### Baseline Confusion Matrix (eDNA Only)

If we applied the `< 45%` rule retrospectively to all fully matched reef/years, these are the outcomes:

```{r confusion-matrix}
matrix_data <- full_matched_data %>%
    mutate(
        Predict_Close = eDNA_Pct_Positive < 45,
        Truth_Needed = (Proportion_NeededLow > 0) | (Proportion_NeededHigh > 0), # Was there an outbreak?
        Confusion_Class = case_when(
            Predict_Close & !Truth_Needed ~ "True Negative (Correctly Closed)",
            Predict_Close & Truth_Needed ~ "False Negative (Missed Outbreak)",
            !Predict_Close & Truth_Needed ~ "True Positive (Correctly Kept Open)",
            !Predict_Close & !Truth_Needed ~ "False Positive (Ghost Culled)"
        )
    )

conf_summary <- matrix_data %>%
    group_by(Confusion_Class) %>%
    summarise(
        `Reef Occurrences` = n(),
        `Diver Hrs Rep` = round(sum(Total_Diver_Hours), 1),
        `0-0.02 Hrs (%)` = paste0(round((sum(Wasted_Diver_Hours) / max(sum(Total_Diver_Hours), 1)) * 100, 1), "%"),
        `0.02-0.04 Hrs (%)` = paste0(round((sum(Borderline_Diver_Hours) / max(sum(Total_Diver_Hours), 1)) * 100, 1), "%"),
        `0.04-0.08 Hrs (%)` = paste0(round((sum(NeededLow_Diver_Hours) / max(sum(Total_Diver_Hours), 1)) * 100, 1), "%"),
        `>0.08 Hrs (%)` = paste0(round((sum(NeededHigh_Diver_Hours) / max(sum(Total_Diver_Hours), 1)) * 100, 1), "%")
    ) %>%
    arrange(desc(`Reef Occurrences`))

kable(conf_summary, caption = "Baseline eDNA Confusion Matrix (with Diver Hour bins)")
```

---

## Multivariate Diagnostic Engine

To aggressively eliminate False Negatives (missed outbreaks) and False Positives (wasted culling trips), we can synthesize additional site data. By bringing in **Manta Tow Scars** and applying an **Initial Scout Cull** (recording CPUE strictly from the first day on-site before committing heavily to the reef), we enhance prediction accuracy.

### Baseline Comparison: Manta Scars Only

To fully understand the value eDNA adds, we must evaluate what happens if we operate solely on traditional Manta Tow scarring data. **Close Reef IF Manta Scars < 10%**:

```{r baseline-scars-only}
matrix_data_scars <- matrix_data %>%
    mutate(
        Pct_Tows_With_Scars = replace_na(Pct_Tows_With_Scars, 0),
        Predict_Close_Scars = (Pct_Tows_With_Scars < 10),
        Confusion_Class_Scars = case_when(
            Predict_Close_Scars & !Truth_Needed ~ "True Negative (Correctly Closed)",
            Predict_Close_Scars & Truth_Needed ~ "False Negative (Missed Outbreak)",
            !Predict_Close_Scars & Truth_Needed ~ "True Positive (Correctly Kept Open)",
            !Predict_Close_Scars & !Truth_Needed ~ "False Positive (Ghost Culled)"
        )
    )

conf_summary_scars <- matrix_data_scars %>%
    group_by(Confusion_Class_Scars) %>%
    summarise(
        `Reef Occurrences` = n(),
        `Diver Hrs Rep` = round(sum(Total_Diver_Hours), 1)
    ) %>%
    arrange(desc(`Reef Occurrences`))

kable(conf_summary_scars, caption = "Univariate Baseline: Scars < 10% (No eDNA factors)")
```

### Hardcoded Rule: eDNA combined with Manta Scars

If we refine the rule to mathematically state: **Close Reef IF (eDNA < 45%) AND (Manta Scars < 10%)**:

```{r mv-hardcoded}
matrix_data_r1 <- matrix_data %>%
    mutate(
        Pct_Tows_With_Scars = replace_na(Pct_Tows_With_Scars, 0),
        Predict_Close = (eDNA_Pct_Positive < 45) & (Pct_Tows_With_Scars < 10),
        Confusion_Class_R1 = case_when(
            Predict_Close & !Truth_Needed ~ "True Negative (Correctly Closed)",
            Predict_Close & Truth_Needed ~ "False Negative (Missed Outbreak)",
            !Predict_Close & Truth_Needed ~ "True Positive (Correctly Kept Open)",
            !Predict_Close & !Truth_Needed ~ "False Positive (Ghost Culled)"
        )
    )

conf_summary_r1 <- matrix_data_r1 %>%
    group_by(Confusion_Class_R1) %>%
    summarise(
        `Reef Occurrences` = n(),
        `Diver Hrs Rep` = round(sum(Total_Diver_Hours), 1)
    ) %>%
    arrange(desc(`Reef Occurrences`))

kable(conf_summary_r1, caption = "Multivariate Rule 1 (eDNA + Scars)")
```

### Baseline Comparison 2: Field Data Only (Scars + Scout Cull)

Before utilizing expensive eDNA processing, what happens if we strictly merge our free, existing field data? **Close Reef IF (Manta Scars < 15%) AND (Initial Scout CPUE < 0.02)**. *(Note: Because this rule inherently requires the Initial Scout Cull, we deduct Scout Diver Hours from the True Negative yield).*

```{r baseline-field-only}
matrix_data_field <- matrix_data %>%
    mutate(
        Pct_Tows_With_Scars = replace_na(Pct_Tows_With_Scars, 0),
        Initial_CPUE = replace_na(Initial_CPUE, 0),
        Predict_Close = (Pct_Tows_With_Scars < 15) & (Initial_CPUE < 0.02),
        Confusion_Class_Field = case_when(
            Predict_Close & !Truth_Needed ~ "True Negative (Correctly Closed)",
            Predict_Close & Truth_Needed ~ "False Negative (Missed Outbreak)",
            !Predict_Close & Truth_Needed ~ "True Positive (Correctly Kept Open)",
            !Predict_Close & !Truth_Needed ~ "False Positive (Ghost Culled)"
        ),
        Net_Saved_Hrs = ifelse(Confusion_Class_Field == "True Negative (Correctly Closed)", Total_Diver_Hours - Scout_Diver_Hours, Total_Diver_Hours)
    )

conf_summary_field <- matrix_data_field %>%
    group_by(Confusion_Class_Field) %>%
    summarise(
        `Reef Occurrences` = n(),
        `Net Diver Hrs Rep` = round(sum(Net_Saved_Hrs, na.rm = TRUE), 1)
    ) %>%
    arrange(desc(`Reef Occurrences`))

kable(conf_summary_field, caption = "Baseline 2: Field Rules strictly (Scars < 15% + Scout Culling < 0.02)")
```

### Hardcoded Rule 2: eDNA, Manta Scars, and Initial Scout CPUE

If we deploy a highly integrated diagnostic expanding the relaxed envelopes but demanding an in-water check: **Close Reef IF (eDNA < 50%) AND (Manta Scars < 15%) AND (Initial Scout CPUE < 0.02)**:
*(Note: Because this rule requires the Initial Scout Cull, we mathematically deduct the Scout Diver Hours from the functionally Saved Hours yield).*

```{r mv-hardcoded-2}
matrix_data_r2 <- matrix_data %>%
    mutate(
        Pct_Tows_With_Scars = replace_na(Pct_Tows_With_Scars, 0),
        Initial_CPUE = replace_na(Initial_CPUE, 0),
        Predict_Close = (eDNA_Pct_Positive < 50) & (Pct_Tows_With_Scars < 15) & (Initial_CPUE < 0.02),
        Confusion_Class_R2 = case_when(
            Predict_Close & !Truth_Needed ~ "True Negative (Correctly Closed)",
            Predict_Close & Truth_Needed ~ "False Negative (Missed Outbreak)",
            !Predict_Close & Truth_Needed ~ "True Positive (Correctly Kept Open)",
            !Predict_Close & !Truth_Needed ~ "False Positive (Ghost Culled)"
        ),
        Net_Saved_Hrs = ifelse(Confusion_Class_R2 == "True Negative (Correctly Closed)", Total_Diver_Hours - Scout_Diver_Hours, Total_Diver_Hours)
    )

conf_summary_r2 <- matrix_data_r2 %>%
    group_by(Confusion_Class_R2) %>%
    summarise(
        `Reef Occurrences` = n(),
        `Net Diver Hrs Rep` = round(sum(Net_Saved_Hrs, na.rm = TRUE), 1)
    ) %>%
    arrange(desc(`Reef Occurrences`))

kable(conf_summary_r2, caption = "Multivariate Rule 2 (eDNA + Scars + Scout Culling)")
```

### Algorithmic Optimization (Decision Trees via rpart)

A decision tree (`rpart`) mathematically scans historical records across eDNA, Manta Tow, and Initial CPUE variables to determine absolute optimal thresholds for closure decisions.
*(Note: Because these models inherently require divers doing an initial scout cull to get `Initial_CPUE`, the ultimate "Saved Hours" in these models formally subtracts those day-1 Scout Diver Hours from the savings yield).*

Instead of blindly eliminating all False Positives or False Negatives completely, we present three models based on standard statistical weighting metrics:

```{r ml-data-prep}
ml_data <- matrix_data %>%
    mutate(
        Pct_Tows_With_Scars = replace_na(Pct_Tows_With_Scars, 0),
        Initial_CPUE = replace_na(Initial_CPUE, 0),
        Truth_Needed_Factor = factor(ifelse(Truth_Needed, "Outbreak", "Ghost"))
    )
```

#### Model 1: Balanced Accuracy

**Focus:** Treats Ghost Culling and Missed Outbreaks equally (1:1 ratio). Optimizes purely for the highest raw predictive accuracy (Gini index). This represents an operational baseline.

```{r ml-balanced}
tree_bal <- rpart(Truth_Needed_Factor ~ eDNA_Pct_Positive + Pct_Tows_With_Scars + Initial_CPUE,
    data = ml_data, method = "class"
)

# Determine if Initial_CPUE was used to conditionally subtract scout hours
used_scout_bal <- "Initial_CPUE" %in% tree_bal$frame$var

ml_data$Pred_Bal <- predict(tree_bal, type = "class")
ml_data <- ml_data %>%
    mutate(
        Class_Bal = case_when(
            Pred_Bal == "Ghost" & Truth_Needed_Factor == "Ghost" ~ "True Negative (Correctly Closed)",
            Pred_Bal == "Ghost" & Truth_Needed_Factor == "Outbreak" ~ "False Negative (Missed Outbreak)",
            Pred_Bal == "Outbreak" & Truth_Needed_Factor == "Outbreak" ~ "True Positive (Correctly Kept Open)",
            Pred_Bal == "Outbreak" & Truth_Needed_Factor == "Ghost" ~ "False Positive (Ghost Culled)"
        ),
        Net_Saved_Bal = ifelse(Class_Bal == "True Negative (Correctly Closed)" & used_scout_bal, Total_Diver_Hours - Scout_Diver_Hours, Total_Diver_Hours)
    )

kable(ml_data %>% group_by(Class_Bal) %>% summarise(`Reef Occurrences` = n(), `Net Diver Hrs` = round(sum(Net_Saved_Bal, na.rm = TRUE), 1)), caption = "Model 1: Balanced Accuracy Outcomes")

plot(tree_bal, margin = 0.1, main = "Threshold Splitting Rules: Balanced Accuracy")
text(tree_bal, use.n = TRUE, all = TRUE, cex = 0.8)
```

#### Model 2: F1-Score Proxy (Harmonic Mean)

**Focus:** The F1 Score acts as the harmonic mean between Precision and Recall. Because there are fewer real outbreaks than ghost reefs (class imbalance), we apply a calculated 2.5-to-1 penalty for missing an outbreak to achieve mathematically resilient F1 classification.

```{r ml-f1}
loss_f1 <- matrix(c(0, 2.5, 1, 0), byrow = TRUE, nrow = 2)
tree_f1 <- rpart(Truth_Needed_Factor ~ eDNA_Pct_Positive + Pct_Tows_With_Scars + Initial_CPUE,
    data = ml_data, parms = list(loss = loss_f1)
)

used_scout_f1 <- "Initial_CPUE" %in% tree_f1$frame$var

ml_data$Pred_F1 <- predict(tree_f1, type = "class")
ml_data <- ml_data %>%
    mutate(
        Class_F1 = case_when(
            Pred_F1 == "Ghost" & Truth_Needed_Factor == "Ghost" ~ "True Negative (Correctly Closed)",
            Pred_F1 == "Ghost" & Truth_Needed_Factor == "Outbreak" ~ "False Negative (Missed Outbreak)",
            Pred_F1 == "Outbreak" & Truth_Needed_Factor == "Outbreak" ~ "True Positive (Correctly Kept Open)",
            Pred_F1 == "Outbreak" & Truth_Needed_Factor == "Ghost" ~ "False Positive (Ghost Culled)"
        ),
        Net_Saved_F1 = ifelse(Class_F1 == "True Negative (Correctly Closed)" & used_scout_f1, Total_Diver_Hours - Scout_Diver_Hours, Total_Diver_Hours)
    )

kable(ml_data %>% group_by(Class_F1) %>% summarise(`Reef Occurrences` = n(), `Net Diver Hrs` = round(sum(Net_Saved_F1, na.rm = TRUE), 1)), caption = "Model 2: F1-Score Harmonic Mean Outcomes")

plot(tree_f1, margin = 0.1, main = "Threshold Splitting Rules: F1-Score")
text(tree_f1, use.n = TRUE, all = TRUE, cex = 0.8)
```

#### Model 3: F2-Score Proxy (Eco-Conservative)

**Focus:** The F2 Score weights Recall (Outbreak Detection) twice as heavily as Precision. We reflect this by imposing a 5-to-1 penalty against False Negatives. This guarantees an ecologically safe approach without collapsing into absolute zero-risk rigidity.

```{r ml-f2}
loss_f2 <- matrix(c(0, 5, 1, 0), byrow = TRUE, nrow = 2)
tree_f2 <- rpart(Truth_Needed_Factor ~ eDNA_Pct_Positive + Pct_Tows_With_Scars + Initial_CPUE,
    data = ml_data, parms = list(loss = loss_f2)
)

used_scout_f2 <- "Initial_CPUE" %in% tree_f2$frame$var

ml_data$Pred_F2 <- predict(tree_f2, type = "class")
ml_data <- ml_data %>%
    mutate(
        Class_F2 = case_when(
            Pred_F2 == "Ghost" & Truth_Needed_Factor == "Ghost" ~ "True Negative (Correctly Closed)",
            Pred_F2 == "Ghost" & Truth_Needed_Factor == "Outbreak" ~ "False Negative (Missed Outbreak)",
            Pred_F2 == "Outbreak" & Truth_Needed_Factor == "Outbreak" ~ "True Positive (Correctly Kept Open)",
            Pred_F2 == "Outbreak" & Truth_Needed_Factor == "Ghost" ~ "False Positive (Ghost Culled)"
        ),
        Net_Saved_F2 = ifelse(Class_F2 == "True Negative (Correctly Closed)" & used_scout_f2, Total_Diver_Hours - Scout_Diver_Hours, Total_Diver_Hours)
    )

kable(ml_data %>% group_by(Class_F2) %>% summarise(`Reef Occurrences` = n(), `Net Diver Hrs` = round(sum(Net_Saved_F2, na.rm = TRUE), 1)), caption = "Model 3: F2-Score Eco-Conservative Outcomes")

plot(tree_f2, margin = 0.1, main = "Threshold Splitting Rules: F2-Score")
text(tree_f2, use.n = TRUE, all = TRUE, cex = 0.8)
```

### Potential Scale vs Operational Program

By converting 20 historical monitoring reefs linearly into tactical targeted reefs, your operational program covers `r operational_tactical_reefs` total closure candidates annually.

For maximum financial transparency, we will directly compare deploying **Approach A (Field Data Only: Scars < 15%, CPUE < 0.02)** which carries zero equipment fees, directly against **Approach B (Highly Integrated Multivariate Rule 2: eDNA < 50%, Scars < 15%, CPUE < 0.02)**:

```{r potential-scale}
yearly_reef_counts <- matrix_data %>%
    group_by(Year) %>%
    summarise(
        `Reefs Tested` = n()
    )

kable(yearly_reef_counts, caption = "Number of Unique Historical Target Validations per Year")
```

```{r matrix-metrics}
Total_Reefs_Ever_Tested <- sum(matrix_data_r2 %>% nrow())

# Approach A: Field Data Only
TN_hrs_field <- sum(matrix_data_field$Net_Saved_Hrs[matrix_data_field$Confusion_Class_Field == "True Negative (Correctly Closed)"], na.rm = TRUE)
FN_count_field <- sum(matrix_data_field$Confusion_Class_Field == "False Negative (Missed Outbreak)")
Avg_Saved_Field <- TN_hrs_field / Total_Reefs_Ever_Tested
Avg_FN_Field <- FN_count_field / Total_Reefs_Ever_Tested
Operation_Saved_Field <- Avg_Saved_Field * operational_tactical_reefs
Operation_FN_Field <- Avg_FN_Field * operational_tactical_reefs
financial_savings_field <- Operation_Saved_Field * cost_per_diver_hour

# Approach B: Field Data + eDNA (Rule 2)
TN_hrs_r2 <- sum(matrix_data_r2$Net_Saved_Hrs[matrix_data_r2$Confusion_Class_R2 == "True Negative (Correctly Closed)"], na.rm = TRUE)
FN_count_r2 <- sum(matrix_data_r2$Confusion_Class_R2 == "False Negative (Missed Outbreak)")
Avg_Saved_R2 <- TN_hrs_r2 / Total_Reefs_Ever_Tested
Avg_FN_R2 <- FN_count_r2 / Total_Reefs_Ever_Tested
Operation_Saved_R2 <- Avg_Saved_R2 * operational_tactical_reefs
Operation_FN_R2 <- Avg_FN_R2 * operational_tactical_reefs
financial_savings_r2 <- Operation_Saved_R2 * cost_per_diver_hour

# Calculate Absolute NET eDNA ROI
cost_edna_testing <- operational_tactical_reefs * edna_cost_per_reef
net_savings_r2 <- financial_savings_r2 - cost_edna_testing
```

#### Approach A: Field Data Only (Scout Culls & Manta Tows)
*   **Operating Cost:** $0 extra.
*   **Hours Safely Saved:** `r format_format(big.mark=",")(round(Operation_Saved_Field, 1))` diver hours returned programmatically per year.
*   **Gross Dollar Savings:** `r dollar(financial_savings_field)`
*   **Operational Risk:** Commits to returning those gross funds at the cost of mathematically missing **`r round(Operation_FN_Field, 1)`** genuine outbreaks per year.

#### Approach B: Integrated eDNA Testing
*   **Operating Cost:** Evaluating `r operational_tactical_reefs` tactical reefs functionally costs **`r dollar(cost_edna_testing)`** annually.
*   **Hours Safely Saved:** Safely recovers `r format_format(big.mark=",")(round(Operation_Saved_R2, 1))` diver hours programmatically per year.
*   **Gross Dollar Savings:** `r dollar(financial_savings_r2)`
*   **Absolute NET Savings:** After paying the $`r format_format(big.mark=",")(cost_edna_testing)` lab/collection fees entirely strictly out of the vessel returns, the net organizational saving clears **`r dollar(net_savings_r2)`**.
*   **Operational Risk:** Bounds the financial returns against an ecologically tighter constraint of missing only **`r round(Operation_FN_R2, 1)`** true localized outbreaks comparatively.

---

## Deep Dive: Most Recent Example Year

The table below provides transparency out of the most recent example year to show you the hour breakdown by precise density bins alongside their associated metrics enabling multivariate closures.

```{r table-recent-year}
# We keep 2025 as the example deep dive
most_recent_year <- max(combined_data$Year - 1, na.rm = TRUE)

recent_top_20 <- top_20_annual %>%
    filter(Year == most_recent_year) %>%
    select(
        ReefName,
        Total_Diver_Hours,
        Wasted_Diver_Hours,
        Proportion_Wasted,
        Proportion_Borderline,
        Proportion_NeededLow,
        Proportion_NeededHigh,
        eDNA_Pct_Positive,
        eDNA_Mean_Conc
    ) %>%
    mutate(
        Total_Diver_Hours = round(Total_Diver_Hours, 1),
        Wasted_Diver_Hours = round(Wasted_Diver_Hours, 1),
        Proportion_Wasted = paste0(round(Proportion_Wasted * 100, 1), "%"),
        Proportion_Borderline = paste0(round(Proportion_Borderline * 100, 1), "%"),
        Proportion_NeededLow = paste0(round(Proportion_NeededLow * 100, 1), "%"),
        Proportion_NeededHigh = paste0(round(Proportion_NeededHigh * 100, 1), "%"),
        eDNA_Pct_Positive = paste0(round(eDNA_Pct_Positive, 1), "%"),
        eDNA_Pct_Positive = ifelse(eDNA_Pct_Positive == "NA%", "No Data", eDNA_Pct_Positive),
        eDNA_Mean_Conc = round(eDNA_Mean_Conc, 3)
    ) %>%
    arrange(desc(Wasted_Diver_Hours))

kable(
    recent_top_20,
    col.names = c("Reef Name", "Tot Dive Hrs", "Wasted Dive Hrs", "% 0-0.02", "% 0.02-0.04", "% 0.04-0.08", "% >0.08", "eDNA % Pos", "eDNA Mean Conc"),
    caption = paste("Worst-Performing Tactical Reefs in", most_recent_year, "(Deep Dive Example)")
)
```

### Spatial Output: Risk Classification Map

To fully visualize how these operational recommendations fall out across the GBR, this interactive map projects every historical validation event natively against the field landscape. Because each Reef may be tested in multiple different years, you can natively toggle the analytical projection overlay by **Year** using the layers control widget on the right.

You can cross-reference our operational "Approach A" (Field Data Only logic) directly against "Approach B" (Integrated eDNA framework) classifications natively over the raw historical sampling footprints.

```{r map-generation}
# 1. Base Geometry Linkage =======================================
edna_raw <- readxl::read_excel("../data/eDNA data_ALL_20260225.xlsx", sheet = "eDNA_data_ALL")
cull_raw <- readxl::read_excel("../data/260201_COTS-Cull-Data-Ewels.xlsx", sheet = "Cull")
manta_raw <- read.csv("../data/COTS Program  Manta Tow Data-2026-02-04.csv", stringsAsFactors = FALSE)

# Generate average lat/long per reef from edna raw data
reef_coords <- edna_raw %>%
    filter(!is.na(Lat) & !is.na(Long)) %>%
    group_by(ReefName = ReefName) %>%
    summarise(
        Lat = mean(as.numeric(Lat), na.rm = TRUE),
        Long = mean(as.numeric(Long), na.rm = TRUE),
        .groups = "drop"
    )

# Bind geometry to our modeled matrices
map_data_r2 <- matrix_data_r2 %>% left_join(reef_coords, by = "ReefName")
map_data_field <- matrix_data_field %>% left_join(reef_coords, by = "ReefName")

# 2. Build Tooltips ============================================
conf_colors <- c(
    "True Negative (Correctly Closed)" = "green",
    "False Negative (Missed Outbreak)" = "red",
    "True Positive (Correctly Kept Open)" = "blue",
    "False Positive (Ghost Culled)" = "gray"
)


# 3. Formulate Background Geo-Layers ============================
# KML Cull Polygons
cull_gpkg <- "../data/Eotr_CotsCullSites_2025_11_19_1_58_PM.gpkg"
if (file.exists(cull_gpkg)) {
    cull_sites_sf <- sf::st_read(cull_gpkg, quiet = TRUE)
} else {
    tmp_kml_dir <- tempfile("kmz_")
    dir.create(tmp_kml_dir)
    unzip("../data/Eotr_CotsCullSites_2025_11_19_1_58_PM.kmz", exdir = tmp_kml_dir)
    kml_layers <- sf::st_layers(file.path(tmp_kml_dir, "doc.kml"))$name
    cull_sites_sf <- do.call(rbind, lapply(kml_layers, function(l) {
        sf::st_read(file.path(tmp_kml_dir, "doc.kml"), layer = l, quiet = TRUE)
    })) |> sf::st_zm(drop = TRUE)
    sf::st_write(cull_sites_sf, dsn = cull_gpkg, driver = "GPKG", append = FALSE, quiet = TRUE)
}

unique_years <- sort(unique(map_data_r2$Year))
max_test_year <- max(unique_years, na.rm = TRUE)

cpue_levels <- c("< 0.01 CPUE", "0.01-0.04 CPUE", "0.04-0.08 CPUE", "> 0.08 CPUE")
cpue_colours <- c("#3288BD", "#FEE08B", "#FC8D59", "#D53E4F")
pal_cpue <- colorFactor(palette = cpue_colours, levels = cpue_levels, ordered = TRUE)

# Cull Sites mapped to latest CPUE Categories
cull_site_latest <- cull_raw |>
    filter(!is.na(CullSiteName)) |>
    filter(SurveyDate >= as.Date(paste0(max_test_year, "-01-01"))) |>
    group_by(CullSiteName) |>
    arrange(desc(SurveyDate)) |>
    slice(1) |>
    ungroup() |>
    mutate(
        site_cots = as.numeric(Cohort1) + as.numeric(Cohort2) + as.numeric(Cohort3) + as.numeric(Cohort4),
        site_cots = ifelse(is.na(site_cots), 0, site_cots),
        site_cpue = round(site_cots / as.numeric(Bottomtime), 4),
        cpue_category = case_when(
            site_cpue >= 0.08 ~ "> 0.08 CPUE",
            site_cpue >= 0.04 ~ "0.04-0.08 CPUE",
            site_cpue >= 0.01 ~ "0.01-0.04 CPUE",
            TRUE ~ "< 0.01 CPUE"
        ),
        cpue_category = factor(cpue_category, levels = cpue_levels)
    )

cull_sites_map <- cull_sites_sf |>
    left_join(cull_site_latest, by = c("Name" = "CullSiteName")) |>
    filter(!is.na(site_cpue))

cull_site_popups <- paste0(
    "<b>Cull Site: ", cull_sites_map$Name, "</b><br>",
    "Latest CPUE: ", cull_sites_map$site_cpue, " COTS/min<br>",
    "<b>Category: ", as.character(cull_sites_map$cpue_category), "</b>"
)

# Manta Tracks (Recent/All)
manta_tracks <- manta_raw |>
    mutate(
        SurveyDate = as.Date(substr(SurveyTime, 1, 10)),
        ScarsCount = case_when(
            !is.na(as.numeric(ScarsCount)) ~ as.numeric(ScarsCount),
            tolower(FeedingScarCountRangeCode) == "a" ~ 0,
            tolower(FeedingScarCountRangeCode) == "p" ~ 4,
            tolower(FeedingScarCountRangeCode) == "c" ~ 10,
            TRUE ~ 0
        )
    ) |>
    filter(SurveyDate >= as.Date(paste0(max_test_year, "-01-01"))) |> # Render only the most recent year's tracks to keep mapping lightweight!
    filter(!is.na(StartLatitude), !is.na(EndLatitude), !is.na(StartLongitude), !is.na(EndLongitude)) |>
    rowwise() |>
    mutate(
        geometry = sf::st_sfc(sf::st_linestring(matrix(c(as.numeric(StartLongitude), as.numeric(EndLongitude), as.numeric(StartLatitude), as.numeric(EndLatitude)), ncol = 2)))
    ) |>
    ungroup() |>
    sf::st_as_sf(crs = 4326)

pal_manta_scars <- colorNumeric(palette = "RdYlBu", domain = 0:10, reverse = TRUE)

manta_cots_pts <- manta_tracks |> filter(as.numeric(CrownOfThornsStarfishCount) > 0)
if (nrow(manta_cots_pts) > 0) manta_cots_pts <- sf::st_centroid(manta_cots_pts)

# Dynamic eDNA internal threshold compute matching the actual analysis limits

cb_calc <- combined_data %>%
    filter(!is.na(eDNA_Pct_Positive), !is.na(eDNA_Mean_Conc), !is.na(Initial_CPUE)) %>%
    mutate(
        conc_t = log1p(eDNA_Mean_Conc),
        cpue = Initial_CPUE,
        perc_pos = eDNA_Pct_Positive
    )

perc_grid <- seq(0, 100, by = 5)
conc_grid_t <- seq(min(cb_calc$conc_t, na.rm = TRUE), quantile(cb_calc$conc_t, 0.99, na.rm = TRUE), length.out = 30)

thr_01 <- best_pair(cb_calc, 0.01, perc_grid, conc_grid_t)
thr_04 <- best_pair(cb_calc, 0.04, perc_grid, conc_grid_t)
thr_08 <- best_pair(cb_calc, 0.08, perc_grid, conc_grid_t)

# Raw eDNA sites
edna_sites <- edna_raw %>%
    mutate(date_edna = as.Date(Date)) %>%
    filter(date_edna >= as.Date(paste0(max_test_year, "-01-01"))) |>
    filter(!is.na(Lat) & !is.na(Long)) |>
    mutate(Lat = as.numeric(Lat), Long = as.numeric(Long)) |>
    group_by(Reef = ReefName, Lat, Long) %>%
    summarise(
        date_edna = min(date_edna, na.rm = T),
        conc_mean = mean(as.numeric(Conc_mean), na.rm = T),
        perc_pos = mean(LOD_sample_positive, na.rm = T) * 100,
        n = n(),
        .groups = "drop"
    ) %>%
    mutate(
        conc_t = log1p(conc_mean),
        cpue_category = case_when(
            perc_pos >= thr_08$perc_star & conc_t >= thr_08$conc_t_star ~ "> 0.08 CPUE",
            perc_pos >= thr_04$perc_star & conc_t >= thr_04$conc_t_star ~ "0.04-0.08 CPUE",
            perc_pos >= thr_01$perc_star & conc_t >= thr_01$conc_t_star ~ "0.01-0.04 CPUE",
            TRUE ~ "< 0.01 CPUE"
        ),
        cpue_category = factor(cpue_category, levels = cpue_levels),
        marker_col = case_when(
            as.integer(cpue_category) == 1L ~ "blue",
            as.integer(cpue_category) == 2L ~ "orange",
            as.integer(cpue_category) == 3L ~ "red",
            as.integer(cpue_category) == 4L ~ "darkred",
            TRUE ~ "gray"
        )
    )

edna_site_popups <- paste0(
    "<b>eDNA Site (", edna_sites$Reef, ")</b><br>",
    "Survey Date: ", format(edna_sites$date_edna, "%d %b %Y"), "<br>",
    "<b>% Positive: ", round(edna_sites$perc_pos, 1), "%</b><br>",
    "<b>Mean Conc: ", round(edna_sites$conc_mean, 2), " copies/rxn</b><br>",
    "Samples: ", edna_sites$n, "<br>",
    "<b>Expected CPUE: ", as.character(edna_sites$cpue_category), "</b>"
)

site_icons <- awesomeIcons(icon = "flask", iconColor = "white", library = "fa", markerColor = edna_sites$marker_col)

# 4. Generate Interactive Map Base =============================
map <- leaflet(width = "100%", height = "700px") |>
    addProviderTiles(providers$Esri.WorldImagery, options = providerTileOptions(maxZoom = 18)) |>
    setView(lng = 147, lat = -18, zoom = 6)

overlay_groups_list <- c()

# Add Background Asset Layers
map <- map |>
    addPolygons(
        data = cull_sites_map,
        color = "#ffffff", weight = 1.0,
        fillColor = ~ pal_cpue(cpue_category),
        fillOpacity = 0.5,
        popup = cull_site_popups,
        group = "Background: KML Cull Polygons"
    ) |>
    addAwesomeMarkers(
        data = edna_sites, lng = ~Long, lat = ~Lat, icon = site_icons, popup = edna_site_popups,
        group = "Background: eDNA Drop Points"
    ) |>
    addPolylines(
        data = manta_tracks, color = ~ pal_manta_scars(ScarsCount), weight = 3, opacity = 0.8,
        popup = ~ paste0("Reef: ", ReefLabel, "<br>COTS: ", CrownOfThornsStarfishCount, "<br>Scars: ", ScarsCount),
        group = "Background: Manta Tows"
    ) |>
    addCircleMarkers(
        data = manta_cots_pts, radius = 4, color = "black", weight = 1, fillColor = "red", fillOpacity = 1,
        popup = ~ paste0("<b>COTS Detected!</b><br>Count: ", CrownOfThornsStarfishCount),
        group = "Background: Manta Tows (COTS > 0)"
    )

overlay_groups_list <- c("Background: KML Cull Polygons", "Background: eDNA Drop Points", "Background: Manta Tows", "Background: Manta Tows (COTS > 0)")

# 5. Inject Predictive Layers Loop (by Year) ====================
unique_years <- sort(unique(map_data_r2$Year))

for (yr in unique_years) {
    yr_data_r2 <- map_data_r2 %>%
        filter(Year == yr) %>%
        filter(!is.na(Lat))
    yr_data_fd <- map_data_field %>%
        filter(Year == yr) %>%
        filter(!is.na(Lat))

    layer_name_r2 <- paste0(yr, " - Approach B: Integrated eDNA")
    layer_name_fd <- paste0(yr, " - Approach A: Field Data Only")

    # Inject Rule 2 Layer
    if (nrow(yr_data_r2) > 0) {
        map <- map %>% addCircleMarkers(
            data = yr_data_r2,
            lng = ~Long, lat = ~Lat,
            radius = 11, weight = 2, color = "white", fillOpacity = 0.9,
            fillColor = ~ unname(conf_colors[Confusion_Class_R2]),
            popup = ~ build_popup(ReefName, Year, eDNA_Pct_Positive, Pct_Tows_With_Scars, Initial_CPUE, Truth_Needed, Confusion_Class_R2, "Approach B: Integrated (eDNA < 50, Scars < 15, CPUE < 0.02)"),
            group = layer_name_r2
        )
    }

    # Inject Field Only Layer
    if (nrow(yr_data_fd) > 0) {
        map <- map %>% addCircleMarkers(
            data = yr_data_fd,
            lng = ~Long, lat = ~Lat,
            radius = 8, weight = 2, color = "black", fillOpacity = 0.9,
            fillColor = ~ unname(conf_colors[Confusion_Class_Field]),
            popup = ~ build_popup(ReefName, Year, eDNA_Pct_Positive, Pct_Tows_With_Scars, Initial_CPUE, Truth_Needed, Confusion_Class_Field, "Approach A: Field Only (Scars < 15, CPUE < 0.02)"),
            group = layer_name_fd
        )
    }

    overlay_groups_list <- c(overlay_groups_list, layer_name_r2, layer_name_fd)
}

# 6. Final Control & Assembly ===================================
map <- map |>
    addLegend(
        position = "bottomright",
        colors = unname(conf_colors),
        labels = names(conf_colors),
        title = "Validation Outcome",
        opacity = 1
    ) %>%
    addLayersControl(
        overlayGroups = overlay_groups_list,
        options = layersControlOptions(collapsed = FALSE)
    ) %>%
    hideGroup(overlay_groups_list) %>%
    showGroup(c("Background: KML Cull Polygons", paste0(max(unique_years), " - Approach B: Integrated eDNA")))

map
```
